---
title: "`r paste0('Importing dataset ', params$series_name, ' and Uploading to GXPA')`"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE
    toc_float: FALSE
    highlight: pygment
    code_folding: hide
    theme:
      primary: "#BE2BBB"
      secondary: "#EEE7E7"
      success: "#1dce9b"
      info: "#009fba"
      warning: "#ffac25"
      danger: "#df603a"
params:
  series_name: "My Dataset"
  series_descript: "Description of my dataset"
  series_organism: "human"
  series_expr_units: "RNAseq log2(TPM + 1)"
  series_default_group: "treatment"
  series_default_color: "timepoint"
  series_analyst: "My Name"
  output_dir_path: "~/output_files"
  input_dir_path: "data"
  expr_counts_filename: "expr_matrix.counts.txt"
  expr_tpm_filename: "expr_matrix.tpm.txt"
  do_log2_tpm_p1: TRUE
  sample_meta_filename: "sample_meta.csv"
  sample_primary_key: "sample_name"
  remove_list: "list(VSLABL = 'Control', sample_name = 'HuRef', sample_type = 'UHR')"
  gene_meta_filename: "gene_metadata.csv"  # e.g. can get this from bioMart for bioconductor org packages
  gene_id_column.from_gene_meta_file: "ensembl_id"  # column in gene metadata file with gene ids matching expr matrix
  gene_symbol_column.from_gene_meta_file: "symbol"  # column in gene metadata file with gene symbols
  gene_id_column.from_expr_file: "ens_id"  # column name of gene_id from expr matrix (cannot be rownames)
  calc_mean_sd: TRUE
  process_score_files: TRUE
  load_session: FALSE
---

```{r setup}

library(stringr)
library(dplyr)
library(DT)
library(data.table)
library(readr)
library(purrr)

# Interactive fallback: load params from this RMDs YAML
if (interactive() && !exists("params")) {
  fm <- tryCatch(
    rmarkdown::yaml_front_matter(knitr::current_input()),
    error = function(e) NULL
  )
  if (!is.null(fm) && !is.null(fm$params)) {
    params <- fm$params
  }
}
```


# DEG score file metadata (entered as R code)

You can specify the metadata here for any differential expression analysis
score files you have run on your own. The output from those analyses should 
be in files where:

   * there is one file for each contrast,
   * the file is in the input directory,
   * that the first column has the gene ids matching those in the expression matrix
   * that it has at least two data columns: log fold change and adjusted p-value (FDR)

```{r deg_meta_params}

# build deg_meta one row at a time with explicit column names
deg_meta <- data.frame(
  score_file_name = character(),
  name = character(),
  description = character(),
  logfc_column = character(),
  fdr_column = integer(),
  group1 = character(),
  group2 = character(),
  stringsAsFactors = FALSE
)

if (params$process_score_files) {
  # you can edit these if you have specific thresholds for score types
  # 1st element is the threshold value, the 2nd is the direction of the sort
  score_thresholds <- list(
    logFC = 1,
    FDR = 0.05
  )

  # for each of your DEG files, add this chunk
  # "name" is how it looks in app, except -logfc or -FDR will be added
  deg_meta[nrow(deg_meta) + 1, ] <- list(
    score_filename = "deseq2_stimX_vs_others.txt",
    name = "stimX-vs-others",
    description = "deseq, stimulation column: stim_x vs __all_others__",
    logfc_column = "logFC",
    fdr_column = "adj.P.Val",
    group1 = "stim_x",
    group2 = "__all_others__"
  )

  
}
```

# Report Parameters (and validations)

```{r params}
if (is.na(params$series_name)) {
  stop("params$series_name can not be missing")
}
if (is.na(params$output_dir_path)) {
  stop("params$output_dir_path can not be missing")
}

# make the output dir path if it does not exist
dir.create(params$output_dir_path, recursive = TRUE, showWarnings = FALSE)

params_for_tab <- lapply(params, function(xx) {
  ifelse(is.null(xx), "NULL", xx)
})

dt_tab <- as.list.data.frame(params_for_tab) %>%
  as.data.frame() %>%
  t() %>%
  as_tibble(rownames = "params", .name_repair = make.names) %>%
  rename(value = X)

datatable(dt_tab,
          rownames = FALSE,
          caption = "Report Parameters",
          options = list(
            autoWidth = TRUE,
            paging = FALSE,
            searching = FALSE
          ))

# also show deg_meta if present
if (nrow(deg_meta) > 0) {
  datatable(deg_meta,
            rownames = FALSE,
            caption = "DEG Score File Metadata",
            options = list(
              autoWidth = TRUE,
              paging = FALSE,
              searching = FALSE
            ))
}

# confirm the gene mapping method vars are present
if (is.null(params$gene_meta_filename) ||
      is.null(params$gene_id_column.from_gene_meta_file) ||
      is.null(params$gene_id_column.from_expr_file) ||
      is.null(params$gene_symbol_column.from_gene_meta_file)) {
  stop("For gene mapping, all of gene_meta_filename, ",
       "gene_id_column.from_gene_meta_file, gene_id_column.from_expr_file, ",
       "and gene_symbol_column.from_gene_meta_file must be specified")
}

# make shorter versions of the gene mapping vars
gene_meta_fn <- params$gene_meta_filename
gene_id_col_meta_file <- params$gene_id_column.from_gene_meta_file
gene_symbol_col_meta_file <- params$gene_symbol_column.from_gene_meta_file
gene_id_col_expr_file <- params$gene_id_column.from_expr_file

# make sure gene_id_col_from_gene_meta_file and gene_symbol_col_
# from_gene_meta_file are not the same
if (gene_id_col_meta_file == gene_symbol_col_meta_file) {
  stop("gene_id_col_meta_file and ",
       "gene_symbol_col_meta_file can not be the same")
}

```


# Confirming files exist in folder: `r params$input_dir_path`

```{r query-xpress, message=FALSE}

if (!dir.exists(params$input_dir_path)) {
  # print the current working directory for debugging
  print(paste("Current working directory:", getwd()))
  stop("Input directory does not exist: ", params$input_dir_path)
}

expr_counts_file_path <-
  file.path(params$input_dir_path, params$expr_counts_filename)
expr_tpm_file_path <-
  file.path(params$input_dir_path, params$expr_tpm_filename)
sample_meta_file_path <-
  file.path(params$input_dir_path, params$sample_meta_filename)

print(paste("expr_counts_file_path:", expr_counts_file_path))
print(paste("expr_tpm_file_path:", expr_tpm_file_path))
print(paste("sample_meta_file_path:", sample_meta_file_path))

# make sure the expression matrix file exists
if (!file.exists(expr_counts_file_path)) {
  stop("Expression counts file does not exist: ", expr_counts_file_path)
}

if (!file.exists(expr_tpm_file_path)) {
  stop("Expression TPM file does not exist: ", expr_tpm_file_path)
}

# make sure the samples file exists
if (!file.exists(sample_meta_file_path)) {
  stop("Samples file does not exist: ", sample_meta_file_path)
}

# make sure all the score files exist
if (nrow(deg_meta) > 0) {
  for (i in 1:nrow(deg_meta)) {
    # print(paste("Checking for score file:", deg_meta$score_file_name[i]))
    score_file_path <-
      file.path(params$input_dir_path, deg_meta$score_file_name[i])

    if (!file.exists(score_file_path)) {
      stop("Score file does not exist: ", score_file_path)
    }
  }
}


```

## Sample Metadata

Remove missing or NA columns, remove samples from the remove list, 
and move columns where all values in the
column are the same to the series file.


```{r samples, message=FALSE}

# read the sample metadata
samples <- read.csv(sample_meta_file_path, stringsAsFactors = FALSE)
print(paste("Num cols metadata:", ncol(samples)))
print(paste("Num rows metadata:", nrow(samples)))

# make sure the params$sample_primary_key is in the samples file
if (!(params$sample_primary_key %in% colnames(samples))) {
  stop("Sample primary key column not found in samples file: ",
       params$sample_primary_key)
}

if (!is.null(params$remove_list)) {
  remove_list <- eval(parse(text = params$remove_list))
  print(paste("remove_list:", paste(remove_list, collapse = ",")))
  print(paste(
    "columns present:",
    paste(names(remove_list)[names(remove_list) %in% colnames(samples)],
          collapse = ",")
  ))
  remove_ids <- purrr::imap(remove_list,
                            function(x, idx) {
                              if (any(colnames(samples) == idx)) {
                                samples[[params$sample_primary_key]][samples[[idx]] %in% x]
                              }
                            }) %>%
    unlist()
  print(paste0("Following IDs removed: ",
               paste0(remove_ids, collapse = ", ")))
} else {
  remove_ids <- character()
}

# keeping only desired samples
keep_ids <- setdiff(samples[[params$sample_primary_key]], remove_ids)
samples <- samples[samples[[params$sample_primary_key]] %in% keep_ids, ]

# rename the sample primary key column to "_id" and move to the first column
colnames(samples)[colnames(samples) == params$sample_primary_key] <- "_id"
samples <- samples %>%
  dplyr::select("_id", dplyr::everything())

# Dropping NA columns
samples <- samples %>%
  dplyr::select(where(~ !all(is.na(.x) | .x %in% c("\"NA\"", ""))))

# Moving some vars which are identical in all rows to series file
cols_to_move <- as.logical(purrr::map_dbl(samples, ~ length(unique(.x))) == 1)
if (any(cols_to_move)) {
  move_to_series <- samples[, cols_to_move, drop = FALSE] %>% unique()

  samples <- samples %>% dplyr::select(!all_of(colnames(move_to_series)))

} else {
  move_to_series <- NULL
}

datatable(samples,
          rownames = FALSE,
          caption = paste0("Sample Metadata (", length(remove_ids),
                           " samples removed)"),
          options = list(
            autoWidth = TRUE,
            paging = TRUE,
            searching = TRUE
          ))

# write the sample metadata to the output dir (this is modified from input)
write_csv(samples,
          file = file.path(params$output_dir_path, "expr.samples.csv"))

```


## Expression Data: Read counts and TPM

```{r read-expr-data-counts-tpm}
# read the expression counts
expr_counts_file_path <-
  file.path(params$input_dir_path, params$expr_counts_filename)
expr_counts <-
  read.delim(expr_counts_file_path,
             stringsAsFactors = FALSE, check.names = FALSE)

# read the expression tpm
expr_tpm_file_path <-
  file.path(params$input_dir_path, params$expr_tpm_filename)
expr_tpm <-
  read.delim(expr_tpm_file_path,
             stringsAsFactors = FALSE, check.names = FALSE)

# confirm `gene_id_col_expr_file` column is in expr_counts and expr_tpm
if (!gene_id_col_expr_file %in% colnames(expr_counts)) {
  stop(paste("gene_id_col_expr_file:",
             gene_id_col_expr_file,
             "not found in expr_counts columns"))
}
if (!gene_id_col_expr_file %in% colnames(expr_tpm)) {
  stop(paste("gene_id_col_expr_file:",
             gene_id_col_expr_file,
             "not found in expr_tpm columns"))
}

# make sure colnames of expr_counts and expr_tpm match (inc gene id column)
if (!all(colnames(expr_counts) == colnames(expr_tpm))) {
  stop("Column names of expr_counts and expr_tpm do not match")
}

```

## Gene Metadata (add symbols to expr matrices)

Read the gene metadata file to get gene ids and symbols and add
the symbols to the expr_counts and expr_tpm dataframes.

```{r feat}

# read the gene metadata file
gene_meta_file_path <- file.path(params$input_dir_path, gene_meta_fn)
gene_meta <- read.csv(gene_meta_file_path, stringsAsFactors = FALSE)

# confirm that gene_id_col_meta_file and gene_symbol_col_meta_file exist in gene_meta
if (!gene_id_col_meta_file %in% colnames(gene_meta)) {
  stop(paste("gene_id_col_meta_file:",
             gene_id_col_meta_file,
             "not found in gene_meta columns"))
}
if (!gene_symbol_col_meta_file %in% colnames(gene_meta)) {
  stop(paste("gene_symbol_col_meta_file:",
             gene_symbol_col_meta_file,
             "not found in gene_meta columns"))
}

# make a new dataframe with just gene ids and symbols so the merge is clean
feat <- gene_meta %>%
  dplyr::select(all_of(c(gene_id_col_meta_file,
                         gene_symbol_col_meta_file)))
colnames(feat) <- c("gene_id", "gene_symbol")

# make sure that most gene ids are in expr_counts and expr_tpm
num_missing_gene_ids_from_counts <-
  sum(!feat$gene_id %in% expr_counts[[gene_id_col_expr_file]])
num_missing_gene_ids_from_tpm <-
  sum(!feat$gene_id %in% expr_tpm[[gene_id_col_expr_file]])

print(paste("Number of gene ids in gene metadata missing from expr_counts:",
            num_missing_gene_ids_from_counts))
print(paste("Number of gene ids in gene metadata missing from expr_tpm:",
            num_missing_gene_ids_from_tpm))

if (num_missing_gene_ids_from_counts / nrow(feat) > 0.2) {
  stop("More than 20% of gene ids in metadata are missing from expr_counts")
}
if (num_missing_gene_ids_from_tpm / nrow(feat) > 0.2) {
  stop("More than 20% of gene ids in metadata are missing from expr_tpm")
}

# add gene symbols to expr_counts and expr_tpm via gene ids
# the resultsing table has gene_id_col_expr_file rather than "gene_id"
expr_counts <- expr_counts %>%
  dplyr::left_join(feat,
                   by = setNames("gene_id", gene_id_col_expr_file))

expr_tpm <- expr_tpm %>%
  dplyr::left_join(feat,
                   by = setNames("gene_id", gene_id_col_expr_file))

# be sure to show which samples are missing from expr matrices or samples
# metadata compared to each other
missing_samples_from_counts <-
  samples$`_id`[!samples$`_id` %in% colnames(expr_counts)]

missing_samples_from_tpm <-
  samples$`_id`[!samples$`_id` %in% colnames(expr_tpm)]

expr_sample_ids <- colnames(expr_tpm)[
  !(colnames(expr_tpm) %in% gene_id_col_expr_file)
]
missing_samples_from_meta <-
  samples$`_id`[!samples$`_id` %in% expr_sample_ids]

if (length(missing_samples_from_counts) > 0) {
  print(paste("Some samples in samples metadata are missing from expr_counts:",
             missing_samples_from_counts))
}
if (length(missing_samples_from_tpm) > 0) {
  print(paste("Some samples in samples metadata are missing from expr_tpm:",
              missing_samples_from_tpm))
}
if (length(missing_samples_from_meta) > 0) {
  print(paste("Some samples in expr matrices are missing from samples metadata:",
              missing_samples_from_meta))
}

```

## Switching to symbols and removing duplicates

Sort the expr data tpm from highest to lowest mean TPM, then remove duplicate
gene symbols keeping the one with the highest mean TPM.

```{r to-hugo-and-clean, warning=FALSE}

expr_tpm_symbols <- expr_tpm %>%
  dplyr::arrange(desc(rowMeans(dplyr::select(., -all_of(c(gene_id_col_expr_file, "gene_symbol")))))) %>%
  dplyr::distinct(gene_symbol, .keep_all = TRUE)

gene_ids_to_keep <- expr_tpm_symbols[[gene_id_col_expr_file]]

expr_counts_symbols <- expr_counts %>%
  dplyr::filter(!!sym(gene_id_col_expr_file) %in% gene_ids_to_keep)


```

todo: print number of rows before and after removing duplicates

## Finalizing Expression Dataframes

Make sure the expr matrix columns are in the same order as the samples dataframe.

```{r finalize-expr-order}

# move symbol to the rownames and remove gene id and symbol columns
expr_tpm_final <- expr_tpm_symbols
expr_counts_final <- expr_counts_symbols
rownames(expr_tpm_final) <- expr_tpm_symbols$gene_symbol
rownames(expr_counts_final) <- expr_counts_symbols$gene_symbol

expr_tpm_final <- expr_tpm_final %>%
  dplyr::select(-all_of(c(gene_id_col_expr_file, "gene_symbol")))
expr_counts_final <- expr_counts_final %>%
  dplyr::select(-all_of(c(gene_id_col_expr_file, "gene_symbol")))

if (params$do_log2_tpm_p1) {
  expr_tpm_final <- round(log2(expr_tpm_final + 1), 5)
}

# restrict the expr data to the samples present in the samples dataframe
# and in the same order
expr_tpm_final <- expr_tpm_final[, match(samples$`_id`, colnames(expr_tpm_final))]
expr_counts_final <- expr_counts_final[, match(samples$`_id`, colnames(expr_counts_final))]

# confirm the dimensions match expectations
if (ncol(expr_tpm_final) != nrow(samples)) {
  stop("Number of columns in expr_tpm_final does not match number of rows in samples")
}

if (ncol(expr_counts_final) != nrow(samples)) {
  stop("Number of columns in expr_counts_final does not match number of rows in samples")
}

# print a few rows
datatable(expr_tpm_final[1:10, 1:min(5,ncol(expr_tpm_final))], 
          rownames = TRUE,
          caption = paste0(
            "Expression Data (tpm): (showing subset; full dataset dim: ",
            nrow(expr_tpm_final)," rows by " , 
            ncol(expr_tpm_final), " columns)"
          ),
          options = list(
            autoWidth = TRUE, 
            paging = FALSE,
            searching = FALSE
          ))

datatable(expr_counts_final[1:10, 1:min(5,ncol(expr_counts_final))], 
          rownames = TRUE,
          caption = paste0(
            "Expression Data (counts): (showing subset; full dataset dim: ",
            nrow(expr_counts_final)," rows by " , 
            ncol(expr_counts_final), " columns)"
          ),
          options = list(
            autoWidth = TRUE, 
            paging = FALSE,
            searching = FALSE
          ))

# write the files to output dir for uploading soon
write.table(expr_tpm_final,
            file = file.path(params$output_dir_path, "expr.expr.txt"),
            sep = "\t", row.names = TRUE, quote = FALSE)

write.table(expr_counts_final,
            file = file.path(params$output_dir_path, "expr.expr_counts.txt"),
            sep = "\t", row.names = TRUE, quote = FALSE)


```

## Reading and modifying DEG score files

After reading the files, keep the ones matching the gene ids in the expression
matrix, and add the rank columns

```{r read-score-files}

# initialize score_info_file dataframe
score_info_file <- data.frame(
  name = character(),
  filename = character(),
  description = character(),
  score_type = character(),
  rank_threshold = integer(),
  score_col = integer(),
  rank_col = integer(),
  group1 = character(),
  group2 = character(),
  stringsAsFactors = FALSE
)

# loop through each score file and read it in
if (nrow(deg_meta) > 0) {
  for (i in 1:nrow(deg_meta)) {
    # print(paste("Checking for score file:", deg_meta$score_file_name[i]))
    score_file_path <-
      file.path(params$input_dir_path, deg_meta$score_file_name[i])

    if (!file.exists(score_file_path)) {
      stop("Score file does not exist: ", score_file_path)
    }

    # read in the score file
    score_data <- read.delim(score_file_path,
                             stringsAsFactors = FALSE,
                             check.names = FALSE)

    # assume first column is gene ids
    colnames(score_data)[1] <- "gene_id"
    
    frac_found <- sum(gene_ids_to_keep %in% score_data$gene_id)/nrow(score_data)
    if (frac_found < 0.2) {
      print(score_file_path)
      stop(paste("many expected gene_ids missing in score file, are the gene_ids in the first column?"))
    }

    # keep only rows matching gene ids in expr_tpm_symbols
    score_data <- score_data %>%
      dplyr::filter(!!sym("gene_id") %in% gene_ids_to_keep)

    # add rank columns for logFC and FDR
    score_data <- score_data %>%
      dplyr::mutate(
        logFC_rank = rank(-abs(.data[[deg_meta$logfc_column[i]]]),
                          ties.method = "first", na.last = "keep"),
        FDR_rank   = rank(.data[[deg_meta$fdr_column[i]]],
                          ties.method = "first", na.last = "keep")
      )

    # add gene symbols to score_data
    score_data <- score_data %>%
      dplyr::left_join(
        feat,
        by = setNames("gene_id", "gene_id")
      )

    # move gene_symbol column to the rownames and remove the gene_symbol column
    rownames(score_data) <- score_data$gene_symbol
    score_data <- score_data %>%
      dplyr::select(-gene_symbol)

    # get the column indices of logFC, FDR, and their ranks
    logfc_col_idx <- which(colnames(score_data) == deg_meta$logfc_column[i])
    fdr_col_idx <- which(colnames(score_data) == deg_meta$fdr_column[i])
    logfc_rank_col_idx <- which(colnames(score_data) == "logFC_rank")
    fdr_rank_col_idx <- which(colnames(score_data) == "FDR_rank")

    logfc_info_row <- data.frame(
      name = paste0(deg_meta$name[i], "-logFC"),
      filename = deg_meta$score_file_name[i],
      description = paste0(deg_meta$description[i], " (logFC)"),
      score_type = "logFC",
      rank_threshold = min(sum(abs(score_data[[deg_meta$logfc_column[i]]]) >= score_thresholds$logFC), 5),
      score_col = logfc_col_idx,
      rank_col = logfc_rank_col_idx,
      group1 = deg_meta$group1[i],
      group2 = deg_meta$group2[i],
      stringsAsFactors = FALSE
    )

    fdr_info_row <- data.frame(
      name = paste0(deg_meta$name[i], "-FDR"),
      filename = deg_meta$score_file_name[i],
      description = paste0(deg_meta$description[i], " (FDR)"),
      score_type = "FDR",
      rank_threshold = min(sum(score_data[[deg_meta$fdr_column[i]]] <= score_thresholds$FDR), 5),
      score_col = fdr_col_idx,
      rank_col = fdr_rank_col_idx,
      group1 = deg_meta$group1[i],
      group2 = deg_meta$group2[i],
      stringsAsFactors = FALSE
    )

    score_info_file <- rbind(score_info_file, logfc_info_row, fdr_info_row)

    # write the modified score_data to new file
    new_file_path <- file.path(params$output_dir_path, deg_meta$score_file_name[i])
    deg_meta$output_file_path[i] <- new_file_path
    write.table(score_data,
                file = new_file_path,
                sep = "\t", row.names = TRUE, quote = FALSE)

  }
}


```


## Calc mean and stdev, if requested

```{r score-files}

if(params$calc_mean_sd) {

  tpm_mat <- expr_tpm_final

  expr_tpm_means <- data.frame(
    gene_symbol = rownames(tpm_mat),
    mean = rowMeans(tpm_mat, na.rm = TRUE),
    stringsAsFactors = FALSE
  ) %>%
    dplyr::arrange(desc(mean)) %>%
    dplyr::mutate(rank = seq_len(dplyr::n()))

  rownames(expr_tpm_means) <- expr_tpm_means$gene_symbol
  expr_tpm_means <- expr_tpm_means %>%
    dplyr::select(-gene_symbol)

  expr_tpm_means$mean <- round(expr_tpm_means$mean, 4)

  write.table(expr_tpm_means, 
              file = file.path(params$output_dir_path, "expr_data.mean.txt"),
              sep = "\t", row.names = TRUE, quote = FALSE)


  expr_tpm_sd <- data.frame(
    gene_symbol = rownames(tpm_mat),
    stdev = apply(tpm_mat, 1, sd, na.rm = TRUE),
    stringsAsFactors = FALSE
  ) %>%
    dplyr::arrange(desc(stdev)) %>%
    dplyr::mutate(rank = seq_len(dplyr::n()))

  rownames(expr_tpm_sd) <- expr_tpm_sd$gene_symbol
  expr_tpm_sd <- expr_tpm_sd %>%
    dplyr::select(-gene_symbol)

  expr_tpm_sd$stdev <- round(expr_tpm_sd$stdev, 4)

  write.table(expr_tpm_sd,
              file = file.path(params$output_dir_path, "expr_data.stdev.txt"),
              sep = "\t", row.names = TRUE, quote = FALSE)

  score_info_file_mean_sd <- data.frame(
    name = c("mean", "stdev"),
    filename = c("expr_data.mean.txt", "expr_data.stdev.txt"),
    description = c("mean across all samples", "stdev across all samples"),
    score_type = c("mean", "stdev"),
    rank_threshold = 200,
    score_col = 1,
    rank_col = 2,
    group1 = "all",
    group2 = ""
  )
  score_info_file <- rbind(score_info_file, score_info_file_mean_sd)
}

# write the score_info_file to output dir
if (nrow(score_info_file) > 0) {
  write.csv(score_info_file,
            file = file.path(params$output_dir_path, "expr.score_info.csv"),
            row.names = FALSE)
}


```


Successfully wrote files to `r params$output_dir_path`

- `r file.path(params$output_dir_path, "expr.samples.csv")`
- `r file.path(params$output_dir_path, "expr.expr.txt")`
- `r file.path(params$output_dir_path, "expr.expr_counts.txt")`
- `r if (params$calc_mean_sd) { file.path(params$output_dir_path, "expr_data.mean.txt") } else { "mean file not generated" }`
- `r if (params$calc_mean_sd) { file.path(params$output_dir_path, "expr_data.stdev.txt") } else { "stdev file not generated" }`
- `r if (nrow(deg_meta) > 0) { paste0(deg_meta$output_file_path, collapse = ", ") } else { "no score files generated" }`
- `r if (nrow(score_info_file) > 0) { file.path(params$output_dir_path, "expr.score_info.csv") } else { "score info file not generated" }`


## Making Series File

Moving any variables with identical values in samples to series file

```{r series-files}


# core series info (as a named list)
series_info <- list(
  series_name = params$series_name,
  series_descript = params$series_descript,
  organism = params$series_organism,
  expr_units = params$series_expr_units,
  default_group = params$series_default_group,
  default_color = params$series_default_color,
  analyst = params$series_analyst
)

# add any constant sample-metadata fields (one row) into the series info
move_to_series_list <- list()
if (!is.null(move_to_series) && nrow(move_to_series) > 0) {
  move_to_series_list <- as.list(move_to_series[1, , drop = TRUE])
}

# combine + coerce values to single strings (safe for CSV key/value)
series_info_all <- c(series_info, move_to_series_list)

series_info_key_val <- data.frame(
  key = names(series_info_all),
  value = vapply(
    series_info_all,
    function(x) paste0(as.character(x), collapse = ";"),
    character(1)
  ),
  stringsAsFactors = FALSE
)

write.csv(series_info_key_val,
          file = file.path(params$output_dir_path, "expr.series_info.csv"),
          row.names = FALSE)

datatable(series_info_key_val,
          rownames = FALSE,
          caption = "Series File",
          options = list(
            autoWidth = TRUE,
            paging = FALSE,
            searching = FALSE
          ))

```

# Everything should be in output_dir, now upload to Needle

You can use the upload_via_interface.Rmd script to upload the files, or you 
can manually upload them via the Needle web interface.

