---
title: "`r paste0('Create Needle files from SE for ', params$series_name)`"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: TRUE
    toc_float: FALSE
    highlight: pygment
    code_folding: hide
params:
  series_name: "My Dataset"
  output_dir_path: "~/output_files"
  input_dir_path: "data"
  se_rdata_filename: "expr_SE.Rdata"
  do_log2_tpm_p1: TRUE
  remove_list: "list(sample_type = 'UHR')"
  gene_symbol_column: "symbol"
  # Optional GXPA metadata
  series_descript: "Description of my dataset"
  series_default_group: "treatment"
  series_default_color: "timepoint"
  series_analyst: "My Name"
  organism: "human"
  expr_units: "RNAseq log2(TPM + 1)"
  process_score_files: TRUE
  calc_mean_sd: TRUE
---

```{r setup}
library(dplyr)
library(readr)
library(purrr)
library(data.table)
library(DT)
library(SummarizedExperiment)
library(GXPAinterface)

# Interactive fallback: load params from this RMDs YAML
if (interactive() && !exists("params")) {
  fm <- tryCatch(
    rmarkdown::yaml_front_matter(knitr::current_input()),
    error = function(e) NULL
  )
  if (!is.null(fm) && !is.null(fm$params)) {
    params <- fm$params
  }
}
```

## Load the saved SummarizedExperiment and switch to Gene Symbols

```{r load-se}
se_path <- file.path(params$output_dir_path, params$se_rdata_filename)
if (!file.exists(se_path)) stop("SE Rdata not found: ", se_path)
load(se_path) # expects object named `se`
if (!exists("se")) stop("Loaded file did not contain object named 'se'")

# apply remove_list if present
if (!is.null(params$remove_list)) {
  remove_list <- eval(parse(text = params$remove_list))
  # Use colData to find IDs to remove
  col_meta_df <- as.data.frame(colData(se))
  remove_ids <- purrr::imap(remove_list, function(x, idx) {
    if (any(colnames(col_meta_df) == idx)) {
      # Use rownames as the sample IDs since _id rename hasn't happened yet
      rownames(col_meta_df)[col_meta_df[[idx]] %in% x]
    }
  }) %>% unlist()
  
  if (length(remove_ids) > 0) {
    se <- se[, !(colnames(se) %in% remove_ids)]
  }
}

# move constant columns into series-level metadata
col_meta_df <- as.data.frame(colData(se))
cols_to_move <- as.logical(purrr::map_dbl(col_meta_df, ~ length(unique(.x))) == 1)
if (any(cols_to_move)) {
  move_to_series <- col_meta_df[, cols_to_move, drop = FALSE] %>% unique()
  # Remove from colData
  colData(se) <- colData(se)[, !cols_to_move, drop = FALSE]
} else {
  move_to_series <- NULL
}

# Switch to symbol: Keep the row with highest mean expression per symbol
row_meta_full <- as.data.frame(rowData(se))
symbol_col <- params$gene_symbol_column
if (!symbol_col %in% colnames(row_meta_full)) stop(paste("rowData(se) must contain symbol column:", symbol_col))

# Create a clear gene_symbol column for downstream code
row_meta_full$gene_symbol <- row_meta_full[[symbol_col]]

# Calculate means to decide which gene_id to keep for each symbol
# Use 'tpm' assay (which is raw from Step 1)
row_means <- rowMeans(assay(se, "tpm"))
row_meta_full$mean_for_dedup <- row_means

best_ids <- row_meta_full %>%
  dplyr::arrange(desc(mean_for_dedup)) %>%
  dplyr::distinct(gene_symbol, .keep_all = TRUE) %>%
  dplyr::pull(gene_id)

se_symbols <- se[best_ids, ]
rownames(se_symbols) <- rowData(se_symbols)[[symbol_col]]

# Now perform log2 transform if requested
if (params$do_log2_tpm_p1) {
  # Save raw for mean/sd later if needed, but the main 'tpm' assay will be log2
  assay(se_symbols, "tpm_raw") <- assay(se_symbols, "tpm")
  assay(se_symbols, "tpm") <- round(log2(assay(se_symbols, "tpm") + 1), 5)
}

# extract matrices and metadata for output
expr_tpm_final <- as.data.frame(assay(se_symbols, "tpm"))
expr_counts_final <- as.data.frame(assay(se_symbols, "counts"))
row_meta <- as.data.frame(rowData(se_symbols))
# Ensure gene_symbol is consistently available in row_meta
row_meta$gene_symbol <- row_meta[[symbol_col]]
col_meta <- as.data.frame(colData(se_symbols))

# ensure `_id` column exists in col_meta (convert rownames to `_id` if necessary)
col_meta_clean <- as.data.frame(col_meta, stringsAsFactors = FALSE)
# remove any existing duplicated id-like columns (e.g., X_id, X_id.1, or existing _id)
col_meta_clean <- col_meta_clean %>% dplyr::select(-dplyr::matches('^X_id(\\.|$)|^_id$'))

# now add a single `_id` column from the rownames as the first column
col_meta_clean <- data.frame(`_id` = rownames(col_meta), col_meta, stringsAsFactors = FALSE, check.names = FALSE)

# remove columns that are identical to the _id column (often the original primary key)
# only remove if it's a character or factor type to be safe
cols_to_check <- which(sapply(col_meta_clean, function(x) is.character(x) || is.factor(x)))
cols_to_check <- setdiff(cols_to_check, 1) # don't check _id against itself

redundant_cols <- sapply(cols_to_check, function(i) {
  all(as.character(col_meta_clean[[i]]) == as.character(col_meta_clean$`_id`))
})

if (any(redundant_cols)) {
  col_meta_clean <- col_meta_clean[, -cols_to_check[redundant_cols], drop = FALSE]
}
col_meta <- col_meta_clean

# force first column name to be `_id` to match expected Needle format
names(col_meta)[1] <- "_id"

# write out sample metadata (colData) as expr.samples.csv
write.table(col_meta, file = file.path(params$output_dir_path, "expr.samples.csv"), sep = ",", row.names = FALSE, quote = FALSE)

# write expression matrices (TPM and counts)
write.table(expr_tpm_final,
            file = file.path(params$output_dir_path, "expr.expr.txt"),
            sep = "\t", row.names = TRUE, quote = FALSE)

write.table(expr_counts_final,
            file = file.path(params$output_dir_path, "expr.expr_counts.txt"),
            sep = "\t", row.names = TRUE, quote = FALSE)
```

## Create series info file

```{r series-info}
# Use any constant columns we moved out of sample metadata as extra info
extra_info <- if (!is.null(move_to_series)) as.list(move_to_series) else NULL

if (is.null(extra_info)) {
  extra_info <- list()
}

extra_info$default_color <- params$series_default_color
extra_info$analyst <- params$series_analyst

GXPAinterface::make_series_info_file(
  filename = file.path(params$output_dir_path, "expr.series_info.csv"),
  series_name = params$series_name,
  series_descript = params$series_descript %||% params$series_name,
  default_group = params$series_default_group,
  organism = params$organism %||% "human",
  expr_units = params$expr_units %||% "TPM",
  extra.list = extra_info
)
```

## DEG score file metadata (enter as R code here)

```{r deg_meta_params}
# build deg_meta one row at a time with explicit column names
deg_meta <- data.frame(
  score_file_name = character(),
  name = character(),
  description = character(),
  logfc_column = character(),
  fdr_column = integer(),
  group1 = character(),
  group2 = character(),
  stringsAsFactors = FALSE
)

if (params$process_score_files) {
  score_thresholds <- list(logFC = 1, FDR = 0.05)

  # example: adjust to your files in params$input_dir_path
  deg_meta[nrow(deg_meta) + 1, ] <- list(
    score_file_name = "deseq2_stimX_vs_others.txt",
    name = "stimX-vs-others",
    description = "deseq, stimulation column: stim_x vs __all_others__",
    logfc_column = "logFC",
    fdr_column = "adj.P.Val",
    group1 = "stim_x",
    group2 = "__all_others__"
  )
}
```

## Read and modify DEG score files (using SE rowData mapping)

```{r read-score-files}
# prepare mapping from gene_id -> gene_symbol using row_meta
if (!("gene_id" %in% colnames(row_meta)) || !("gene_symbol" %in% colnames(row_meta))) {
  # try to recover if gene_id was stored under a different name
  if ("gene_symbol" %in% colnames(row_meta)) {
    row_meta$gene_id <- ifelse(!is.null(row_meta$gene_id), row_meta$gene_id, NA)
  } else {
    stop("rowData(se) must contain at least gene_symbol (and ideally gene_id)")
  }
}

feat <- row_meta %>% dplyr::select(gene_id, gene_symbol)
rownames(feat) <- feat$gene_symbol

# gene ids to keep (from SE)
gene_ids_to_keep <- feat$gene_id

# initialize score_info_file dataframe
score_info_file <- data.frame(
  name = character(),
  filename = character(),
  description = character(),
  score_type = character(),
  rank_threshold = integer(),
  score_col = integer(),
  rank_col = integer(),
  group1 = character(),
  group2 = character(),
  stringsAsFactors = FALSE
)

if (nrow(deg_meta) > 0) {
  for (i in 1:nrow(deg_meta)) {
    score_file_path <- file.path(params$input_dir_path, deg_meta$score_file_name[i])
    if (!file.exists(score_file_path)) stop("Score file does not exist: ", score_file_path)

    score_data <- read.delim(score_file_path, stringsAsFactors = FALSE, check.names = FALSE)
    colnames(score_data)[1] <- "gene_id"

    frac_found <- sum(gene_ids_to_keep %in% score_data$gene_id)/nrow(score_data)
    if (frac_found < 0.2) stop(paste("many expected gene_ids missing in score file, are the gene_ids in the first column?", score_file_path))

    score_data <- score_data %>% dplyr::filter(!!rlang::sym("gene_id") %in% gene_ids_to_keep)

    score_data <- score_data %>% dplyr::mutate(
      logFC_rank = rank(-abs(.data[[deg_meta$logfc_column[i]]]), ties.method = "first", na.last = "keep"),
      FDR_rank   = rank(.data[[deg_meta$fdr_column[i]]], ties.method = "first", na.last = "keep")
    )

    score_data <- score_data %>% dplyr::left_join(feat, by = setNames("gene_id", "gene_id"))
    rownames(score_data) <- score_data$gene_symbol
    score_data <- score_data %>% dplyr::select(-gene_symbol)

    logfc_col_idx <- which(colnames(score_data) == deg_meta$logfc_column[i])
    fdr_col_idx <- which(colnames(score_data) == deg_meta$fdr_column[i])
    logfc_rank_col_idx <- which(colnames(score_data) == "logFC_rank")
    fdr_rank_col_idx <- which(colnames(score_data) == "FDR_rank")

    logfc_info_row <- data.frame(
      name = paste0(deg_meta$name[i], "-logFC"),
      filename = deg_meta$score_file_name[i],
      description = paste0(deg_meta$description[i], " (logFC)"),
      score_type = "logFC",
      rank_threshold = min(sum(abs(score_data[[deg_meta$logfc_column[i]]]) >= score_thresholds$logFC), 5),
      score_col = logfc_col_idx,
      rank_col = logfc_rank_col_idx,
      group1 = deg_meta$group1[i],
      group2 = deg_meta$group2[i],
      stringsAsFactors = FALSE
    )

    fdr_info_row <- data.frame(
      name = paste0(deg_meta$name[i], "-FDR"),
      filename = deg_meta$score_file_name[i],
      description = paste0(deg_meta$description[i], " (FDR)"),
      score_type = "FDR",
      rank_threshold = min(sum(score_data[[deg_meta$fdr_column[i]]] <= score_thresholds$FDR), 5),
      score_col = fdr_col_idx,
      rank_col = fdr_rank_col_idx,
      group1 = deg_meta$group1[i],
      group2 = deg_meta$group2[i],
      stringsAsFactors = FALSE
    )

    score_info_file <- rbind(score_info_file, logfc_info_row, fdr_info_row)

    new_file_path <- file.path(params$output_dir_path, deg_meta$score_file_name[i])
    deg_meta$output_file_path[i] <- new_file_path
    write.table(score_data, file = new_file_path, sep = "\t", row.names = TRUE, quote = FALSE)
  }
}
```

## Calc mean and stdev, if requested

```{r mean-sd}
if (params$calc_mean_sd) {
  # Use the main TPM assay (which is log2 if do_log2_tpm_p1 was TRUE)
  tpm_mat <- as.data.frame(assay(se_symbols, "tpm"))

  expr_tpm_means <- data.frame(
    gene_symbol = rownames(tpm_mat),
    mean = rowMeans(tpm_mat, na.rm = TRUE),
    stringsAsFactors = FALSE
  ) %>% dplyr::arrange(desc(mean)) %>% dplyr::mutate(rank = seq_len(n()))

  rownames(expr_tpm_means) <- expr_tpm_means$gene_symbol
  expr_tpm_means <- expr_tpm_means %>% dplyr::select(-gene_symbol)
  expr_tpm_means$mean <- round(expr_tpm_means$mean, 4)

  write.table(expr_tpm_means, file = file.path(params$output_dir_path, "expr_data.mean.txt"), sep = "\t", row.names = TRUE, quote = FALSE)

  expr_tpm_sd <- data.frame(
    gene_symbol = rownames(tpm_mat),
    stdev = apply(tpm_mat, 1, sd, na.rm = TRUE),
    stringsAsFactors = FALSE
  ) %>% dplyr::arrange(desc(stdev)) %>% dplyr::mutate(rank = seq_len(n()))

  rownames(expr_tpm_sd) <- expr_tpm_sd$gene_symbol
  expr_tpm_sd <- expr_tpm_sd %>% dplyr::select(-gene_symbol)
  expr_tpm_sd$stdev <- round(expr_tpm_sd$stdev, 4)

  write.table(expr_tpm_sd, file = file.path(params$output_dir_path, "expr_data.stdev.txt"), sep = "\t", row.names = TRUE, quote = FALSE)

  score_info_file_mean_sd <- data.frame(
    name = c("mean", "stdev"),
    filename = c("expr_data.mean.txt", "expr_data.stdev.txt"),
    description = c("mean across all samples", "stdev across all samples"),
    score_type = c("mean", "stdev"),
    rank_threshold = 200,
    score_col = 1,
    rank_col = 2,
    group1 = "all",
    group2 = "",
    stringsAsFactors = FALSE
  )
  score_info_file <- rbind(score_info_file, score_info_file_mean_sd)
}

if (nrow(score_info_file) > 0) {
  write.csv(score_info_file, file = file.path(params$output_dir_path, "expr.score_info.csv"), row.names = FALSE)
}
```

Files written to `r params$output_dir_path`:

- `r file.path(params$output_dir_path, "expr.samples.csv")`
- `r file.path(params$output_dir_path, "expr.series_info.csv")`
- `r file.path(params$output_dir_path, "expr.expr.txt")`
- `r file.path(params$output_dir_path, "expr.expr_counts.txt")`
- `r if (params$calc_mean_sd) { file.path(params$output_dir_path, "expr_data.mean.txt") } else { "mean file not generated" }`
- `r if (params$calc_mean_sd) { file.path(params$output_dir_path, "expr_data.stdev.txt") } else { "stdev file not generated" }`
- `r if (nrow(deg_meta) > 0) { paste0(deg_meta$output_file_path, collapse = ", ") } else { "no score files generated" }`
- `r if (nrow(score_info_file) > 0) { file.path(params$output_dir_path, "expr.score_info.csv") } else { "score info file not generated" }`
