---
title: "`r paste0('Build SummarizedExperiment for ', params$series_name)`"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: TRUE
    toc_float: FALSE
    highlight: pygment
    code_folding: hide
params:
  series_name: "My Dataset"
  output_dir_path: "~/output_files"
  input_dir_path: "data"
  expr_counts_filename: "expr_matrix.counts.txt"
  expr_tpm_filename: "expr_matrix.tpm.txt"
  do_log2_tpm_p1: TRUE
  sample_meta_filename: "sample_meta.csv"
  sample_primary_key: "sample_name"
  remove_list: "list(VSLABL = 'Control', sample_name = 'HuRef', sample_type = 'UHR')"
  gene_meta_filename: "gene_metadata.csv"
  gene_id_column.from_gene_meta_file: "ensembl_id"
  gene_symbol_column.from_gene_meta_file: "symbol"
  gene_id_column.from_expr_file: "ens_id"
---

```{r setup}
library(stringr)
library(dplyr)
library(data.table)
library(readr)
library(purrr)
library(SummarizedExperiment)

# Interactive fallback: load params from this RMDs YAML
if (interactive() && !exists("params")) {
  fm <- tryCatch(
    rmarkdown::yaml_front_matter(knitr::current_input()),
    error = function(e) NULL
  )
  if (!is.null(fm) && !is.null(fm$params)) {
    params <- fm$params
  }
}
```

# Read and validate input files

```{r params}
# basic checks
if (is.na(params$series_name)) stop("params$series_name can not be missing")
if (is.na(params$output_dir_path)) stop("params$output_dir_path can not be missing")

dir.create(params$output_dir_path, recursive = TRUE, showWarnings = FALSE)

# shorten gene mapping vars
gene_meta_fn <- params$gene_meta_filename
gene_id_col_meta_file <- params$gene_id_column.from_gene_meta_file
gene_symbol_col_meta_file <- params$gene_symbol_column.from_gene_meta_file
gene_id_col_expr_file <- params$gene_id_column.from_expr_file
```

## Confirm files exist

```{r confirm-files}
input_dir <- params$input_dir_path
expr_counts_file_path <- file.path(input_dir, params$expr_counts_filename)
expr_tpm_file_path <- file.path(input_dir, params$expr_tpm_filename)
sample_meta_file_path <- file.path(input_dir, params$sample_meta_filename)

if (!dir.exists(input_dir)) stop("Input directory does not exist: ", input_dir)
if (!file.exists(expr_counts_file_path)) stop("Expression counts file does not exist: ", expr_counts_file_path)
if (!file.exists(expr_tpm_file_path)) stop("Expression TPM file does not exist: ", expr_tpm_file_path)
if (!file.exists(sample_meta_file_path)) stop("Samples file does not exist: ", sample_meta_file_path)
if (!file.exists(file.path(input_dir, gene_meta_fn))) stop("Gene metadata file not found: ", gene_meta_fn)
```

## Sample metadata cleaning

```{r samples}
# read samples
samples <- read.csv(sample_meta_file_path, stringsAsFactors = FALSE)
if (!(params$sample_primary_key %in% colnames(samples))) stop("Sample primary key column not found in samples file: ", params$sample_primary_key)

# apply remove_list if present
if (!is.null(params$remove_list)) {
  remove_list <- eval(parse(text = params$remove_list))
  remove_ids <- purrr::imap(remove_list, function(x, idx) {
    if (any(colnames(samples) == idx)) {
      samples[[params$sample_primary_key]][samples[[idx]] %in% x]
    }
  }) %>% unlist()
} else {
  remove_ids <- character()
}

keep_ids <- setdiff(samples[[params$sample_primary_key]], remove_ids)
samples <- samples[samples[[params$sample_primary_key]] %in% keep_ids, ]

# rename primary key to _id and move first
colnames(samples)[colnames(samples) == params$sample_primary_key] <- "_id"
samples <- samples %>% dplyr::select("_id", dplyr::everything())

# drop empty/NA columns
samples <- samples %>% dplyr::select(where(~ !all(is.na(.x) | .x %in% c("\"NA\"", ""))))

# move constant columns into series-level metadata
cols_to_move <- as.logical(purrr::map_dbl(samples, ~ length(unique(.x))) == 1)
if (any(cols_to_move)) {
  move_to_series <- samples[, cols_to_move, drop = FALSE] %>% unique()
  samples <- samples %>% dplyr::select(!all_of(colnames(move_to_series)))
} else {
  move_to_series <- NULL
}

# write a samples file (modified) for reference
write_csv(samples, file = file.path(params$output_dir_path, "expr.samples.csv"))
```

## Read expression matrices

```{r read-expr}
expr_counts <- read.delim(expr_counts_file_path, stringsAsFactors = FALSE, check.names = FALSE)
expr_tpm <- read.delim(expr_tpm_file_path, stringsAsFactors = FALSE, check.names = FALSE)

# ensure gene id column present
if (!gene_id_col_expr_file %in% colnames(expr_counts)) stop(paste("gene_id_col_expr_file:", gene_id_col_expr_file, "not found in expr_counts columns"))
if (!gene_id_col_expr_file %in% colnames(expr_tpm)) stop(paste("gene_id_col_expr_file:", gene_id_col_expr_file, "not found in expr_tpm columns"))

if (!all(colnames(expr_counts) == colnames(expr_tpm))) stop("Column names of expr_counts and expr_tpm do not match")
```

## Gene metadata and adding symbols

```{r feat}
gene_meta <- read.csv(file.path(input_dir, gene_meta_fn), stringsAsFactors = FALSE)
if (!gene_id_col_meta_file %in% colnames(gene_meta)) stop(paste("gene_id_col_meta_file:", gene_id_col_meta_file, "not found in gene_meta columns"))
if (!gene_symbol_col_meta_file %in% colnames(gene_meta)) stop(paste("gene_symbol_col_meta_file:", gene_symbol_col_meta_file, "not found in gene_meta columns"))

feat <- gene_meta %>% dplyr::select(all_of(c(gene_id_col_meta_file, gene_symbol_col_meta_file)))
colnames(feat) <- c("gene_id", "gene_symbol")

# check mapping coverage
num_missing_gene_ids_from_counts <- sum(!feat$gene_id %in% expr_counts[[gene_id_col_expr_file]])
num_missing_gene_ids_from_tpm <- sum(!feat$gene_id %in% expr_tpm[[gene_id_col_expr_file]])

if (num_missing_gene_ids_from_counts / nrow(feat) > 0.2) stop("More than 20% of gene ids in metadata are missing from expr_counts")
if (num_missing_gene_ids_from_tpm / nrow(feat) > 0.2) stop("More than 20% of gene ids in metadata are missing from expr_tpm")

# join symbols into expression tables
expr_counts <- expr_counts %>% dplyr::left_join(feat, by = setNames("gene_id", gene_id_col_expr_file))
expr_tpm <- expr_tpm %>% dplyr::left_join(feat, by = setNames("gene_id", gene_id_col_expr_file))
```

## Keep highest-TPM symbol per gene and align with samples

```{r dedup}
expr_tpm_symbols <- expr_tpm %>%
  dplyr::arrange(desc(rowMeans(dplyr::select(., -all_of(c(gene_id_col_expr_file, "gene_symbol")))))) %>%
  dplyr::distinct(gene_symbol, .keep_all = TRUE)

gene_ids_to_keep <- expr_tpm_symbols[[gene_id_col_expr_file]]

expr_counts_symbols <- expr_counts %>% dplyr::filter(!!rlang::sym(gene_id_col_expr_file) %in% gene_ids_to_keep)

# move symbol to rownames and drop id/symbol columns
expr_tpm_final <- expr_tpm_symbols
expr_counts_final <- expr_counts_symbols
rownames(expr_tpm_final) <- expr_tpm_symbols$gene_symbol
rownames(expr_counts_final) <- expr_counts_symbols$gene_symbol
expr_tpm_final <- expr_tpm_final %>% dplyr::select(-all_of(c(gene_id_col_expr_file, "gene_symbol")))
expr_counts_final <- expr_counts_final %>% dplyr::select(-all_of(c(gene_id_col_expr_file, "gene_symbol")))

if (params$do_log2_tpm_p1) {
  expr_tpm_final <- round(log2(expr_tpm_final + 1), 5)
}

# preserve raw TPM prior to any transform for mean/stdev calculations
tpm_raw <- expr_tpm_symbols %>%
  dplyr::select(-all_of(c(gene_id_col_expr_file, "gene_symbol")))

# if we applied log transform above, `expr_tpm_final` holds transformed values
# otherwise it equals `tpm_raw` already

# restrict to samples present and match order
expr_tpm_final <- expr_tpm_final[, match(samples$`_id`, colnames(expr_tpm_final))]
expr_counts_final <- expr_counts_final[, match(samples$`_id`, colnames(expr_counts_final))]

# align tpm_raw to same row/column structure
rownames(tpm_raw) <- expr_tpm_symbols$gene_symbol
tpm_raw <- tpm_raw[, match(samples$`_id`, colnames(tpm_raw))]

if (ncol(expr_tpm_final) != nrow(samples)) stop("Number of columns in expr_tpm_final does not match number of rows in samples")
if (ncol(expr_counts_final) != nrow(samples)) stop("Number of columns in expr_counts_final does not match number of rows in samples")
```

## Build SummarizedExperiment and save

```{r build-se}
# prepare rowData and colData
row_df <- data.frame(gene_symbol = rownames(expr_tpm_final), stringsAsFactors = FALSE)
# attempt to preserve original gene_id mapping
row_df$gene_id <- expr_tpm_symbols[[gene_id_col_expr_file]]
rownames(row_df) <- row_df$gene_symbol

col_df <- samples
rownames(col_df) <- col_df$`_id`

if (params$do_log2_tpm_p1) {
  assays_list <- list(
    tpm = as.matrix(expr_tpm_final),    # transformed TPM (log2)
    tpm_raw = as.matrix(tpm_raw),       # raw TPM for mean/stdev
    counts = as.matrix(expr_counts_final)
  )
} else {
  assays_list <- list(
    tpm = as.matrix(expr_tpm_final),    # raw TPM
    counts = as.matrix(expr_counts_final)
  )
}

se <- SummarizedExperiment(assays = assays_list, rowData = row_df, colData = col_df)

save(se, file = file.path(params$output_dir_path, "expr_SE.Rdata"))

message("Saved SummarizedExperiment to: ", file.path(params$output_dir_path, "expr_SE.Rdata"))
```

# Done

Files written to `r params$output_dir_path`:

- `r file.path(params$output_dir_path, "expr_SE.Rdata")`
- `r file.path(params$output_dir_path, "expr.samples.csv")`
