---
title: "`r paste0('Create Needle files from SE for ', params$series_name)`"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: TRUE
    toc_float: FALSE
    highlight: pygment
    code_folding: hide
params:
  series_name: "My Dataset"
  output_dir_path: "~/output_files"
  input_dir_path: "data"
  se_rdata_filename: "expr_SE.Rdata"
  do_log2_tpm_p1: TRUE
  remove_list: "list(sample_type = 'UHR')"
  gene_symbol_column: "symbol"
  # Optional GXPA metadata
  series_descript: "Description of my dataset"
  series_default_group: "treatment"
  series_default_color: "timepoint"
  series_analyst: "My Name"
  organism: "human"
  expr_units: "RNAseq log2(TPM + 1)"
  process_score_files: TRUE
  calc_mean_sd: TRUE
---

```{r setup}
library(dplyr)
library(readr)
library(purrr)
library(data.table)
library(DT)
library(SummarizedExperiment)
library(GXPAinterface)

# Interactive fallback: load params from this RMDs YAML
if (interactive() && !exists("params")) {
  fm <- tryCatch(
    rmarkdown::yaml_front_matter(knitr::current_input()),
    error = function(e) NULL
  )
  if (!is.null(fm) && !is.null(fm$params)) {
    params <- fm$params
  }
}
```

## Load the saved SummarizedExperiment and pre-process

```{r load-se}
se_path <- file.path(params$output_dir_path, params$se_rdata_filename)
if (!file.exists(se_path)) stop("SE Rdata not found: ", se_path)
load(se_path) # expects object named `se`
if (!exists("se")) stop("Loaded file did not contain object named 'se'")

# 1. Filter samples
if (!is.null(params$remove_list)) {
  remove_list <- eval(parse(text = params$remove_list))
  se <- GXPAinterface::filter_se_samples(se, remove_list)
}

# 2. Extract constant metadata for series info
se_list <- GXPAinterface::extract_se_constant_metadata(se)
se <- se_list$se
extra_info <- se_list$constant_meta

# 3. Deduplicate by symbol (keeps row with highest mean expr)
se_symbols <- GXPAinterface::deduplicate_se_by_symbol(
  se = se,
  symbol_col = params$gene_symbol_column,
  primary_assay = if ("tpm" %in% SummarizedExperiment::assayNames(se)) "tpm" else "counts"
)

# 4. Perform log2 transform on TPM if requested
if (params$do_log2_tpm_p1 && "tpm" %in% SummarizedExperiment::assayNames(se_symbols)) {
  assay(se_symbols, "tpm_raw") <- SummarizedExperiment::assay(se_symbols, "tpm")
  SummarizedExperiment::assay(se_symbols, "tpm") <- round(log2(SummarizedExperiment::assay(se_symbols, "tpm") + 1), 5)
}

# 5. Write expression and sample metadata to GXPA flat files
GXPAinterface::write_se_to_gxpa(
  se = se_symbols,
  output_dir = params$output_dir_path,
  run_checks = TRUE
)
```

## Create series info file

```{r series-info}
extra_info$default_color <- params$series_default_color
extra_info$analyst <- params$series_analyst

GXPAinterface::make_series_info_file(
  filename = file.path(params$output_dir_path, "expr.series_info.csv"),
  series_name = params$series_name,
  series_descript = params$series_descript %||% params$series_name,
  default_group = params$series_default_group,
  organism = params$organism %||% "human",
  expr_units = params$expr_units %||% "TPM",
  extra.list = extra_info
)
```

## DEG score file metadata (enter as R code here)

```{r deg_meta_params}
# build deg_meta one row at a time with explicit column names
deg_meta <- data.frame(
  score_file_name = character(),
  name = character(),
  description = character(),
  logfc_column = character(),
  fdr_column = character(),
  group1 = character(),
  group2 = character(),
  stringsAsFactors = FALSE
)

if (params$process_score_files) {
  # example: adjust to your files in params$input_dir_path
  deg_meta[nrow(deg_meta) + 1, ] <- list(
    score_file_name = "deseq2_stimX_vs_others.txt",
    name = "stimX-vs-others",
    description = "deseq, stimulation column: stim_x vs __all_others__",
    logfc_column = "logFC",
    fdr_column = "adj.P.Val",
    group1 = "stim_x",
    group2 = "__all_others__"
  )
}
```

## Read and modify DEG score files (using SE rowData mapping)

```{r read-score-files}
# gene mapping for score processing (maps original unique_gene_id to symbols)
gene_map <- as.data.frame(SummarizedExperiment::rowData(se_symbols)) %>% 
  dplyr::select(gene_id = !!rlang::sym("unique_gene_id"), 
                gene_symbol = !!rlang::sym(params$gene_symbol_column))

score_info_file <- GXPAinterface::process_gxpa_scores(
  deg_meta = deg_meta,
  gene_id_map = gene_map,
  input_dir = params$input_dir_path,
  output_dir = params$output_dir_path,
  score_thresholds = list(logFC = 1, FDR = 0.05)
)
```

## Calc mean and stdev, if requested

```{r mean-sd}
if (params$calc_mean_sd) {
  # Use the main TPM assay
  tpm_mat <- as.matrix(SummarizedExperiment::assay(se_symbols, "tpm"))
  
  mean_sd_info <- GXPAinterface::write_mean_sd_scores(
    tpm_mat = tpm_mat,
    output_dir = params$output_dir_path
  )
  
  score_info_file <- rbind(score_info_file, mean_sd_info)
}

if (nrow(score_info_file) > 0) {
  utils::write.csv(score_info_file, file = file.path(params$output_dir_path, "expr.score_info.csv"), row.names = FALSE)
}
```

Files written to `r params$output_dir_path`:

- `r file.path(params$output_dir_path, "expr.samples.csv")`
- `r file.path(params$output_dir_path, "expr.series_info.csv")`
- `r file.path(params$output_dir_path, "expr.expr.txt")`
- `r file.path(params$output_dir_path, "expr.expr_counts.txt")`
- `r if (params$calc_mean_sd) { file.path(params$output_dir_path, "expr_data.mean.txt") } else { "mean file not generated" }`
- `r if (params$calc_mean_sd) { file.path(params$output_dir_path, "expr_data.stdev.txt") } else { "stdev file not generated" }`
- `r if (nrow(deg_meta) > 0) { paste0(deg_meta$output_file_path, collapse = ", ") } else { "no score files generated" }`
- `r if (nrow(score_info_file) > 0) { file.path(params$output_dir_path, "expr.score_info.csv") } else { "score info file not generated" }`
